Towards an Automated Sportscasting System for Chess

### 2011-05-18 paper/paper.tex
Final paper

\documentclass[11pt]{article}
\usepackage{acl-hlt2011}
\usepackage{color}
\usepackage{graphicx}
\usepackage{times}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{algorithmic}
\usepackage{multirow}
\usepackage{url}
\DeclareMathOperator*{\argmax}{arg\,max}
\setlength\titlebox{6.5cm}    % Expanding the titlebox

@@

\begin{abstract}
We investigate applications of statistical learning to automated sportscasting for chess. In particular, this project studies methods to automatically identify high level structures in chess games using support vector machines and auto-encoders. We then compare the results of the two different approaches and showed improvements over the SVM model using a stacked auto-encoder. Finally, we conclude with results, analysis of errors, and ideas for future work.
\end{abstract}

@@

Unfortunately, we found that our approaches did not deliver accurate results in our experiments. In particular, the auto-encoder's output failed to serve as a useful criterion for distinguishing between different types of themes. We examine the types of correct predictions made by our classifiers and discuss how we may improve our results in the future.

@@

\paragraph{Deep learning.} These techniques attempt to capture complex relationships between input variables in order to transform the input into a different representation which better expresses the relevant structures that make up the input. Specifically, they compose many layers of non-linearities in order to compactly represent complex functions which can model these relationships. They have been used successfully for many different types of tasks \cite{Vincent,Lee,Socher}.

@@

However, we identified some shortcomings of the AMBL system. First, it relies on experts to provide arguments as to why a certain position exemplifies (or doesn’t) a particular positional feature. While this is an improvement from defining a classification rule manually, it is infeasible to repeat this procedure multiple times. Second, the arguments that the experts are allowed to provide rely on a preset list of features.

@@

\begin{table*}
\centering
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{Name} & \textbf{Description} \\ \hline
\tt{attackPossible} & The current player has opportunity to attack. \\
\tt{backrankWeakness} & There exists a future checkmate by a rook or queen on the edges of the board. \\
\tt{bishopPair} & There exists a decisive strategic advantage for the player with two bishops. \\
\tt{clearance} & There exists a maneuver to break the pawn structure. \\ 
\tt{closure} & There exists a maneuver to fix the pawn structure. \\
\tt{decoy} & There exists a maneuver to distract a crucial piece. \\
\tt{defensePressure} & There exists a maneuver to break the opposing King’s defense. \\
\tt{deflection} & There exists a maneuver to deflect an attack.  \\ 
\tt{discoveredAttack} & There exists a maneuver to allow for a discovered attack (an attack that was blocked by a friendly piece.) \\
\tt{doubleAttack} & There exists a move that simultaneously attacks at least two crucial squares. \\
\tt{drawCombinationsStalemate} & There exists a combination that allows for decisive material advantage or immediate checkmate. \\
\tt{enclosedKings} & There exists a maneuver to trap the opposing King. \\
\tt{freeingBlockingSquares} & There exists a move to free a crucial square. \\
\tt{goodEndgameChances} & There exists a move to allow for a favorable endgame. \\
\tt{inBetweenMove} & There exists an interrupting move that allows for decisive material advantage of immediate checkmate. \\ 
\tt{interference} & There exists a maneuver to block a crucial piece. \\ 
\tt{interruption} & There exists a maneuver that interrupts the opponent’s tactics. \\
\tt{kingInCenter} & There exists a decisive strategic advantage for the player with the King in the center of the board. \\
\tt{kingToBadSquares} & There exists a combination that forces the opposing King to move to a bad square. \\
\tt{longDiagonals} & There exists a strategic advantage for the player that controls squares on the long diagonal. \\
\tt{matingAttack} & There exists a decisive mating attack. \\
\tt{openingClosingDiagonals} & There exists a maneuver to clear a diagonal. \\
\tt{openingClosingFiles} & There exists a maneuver to clear a file. \\
\tt{overload} & There exists a combination that makes a defending piece defend too many squares at once. \\
\tt{passedPawn} & There exists a combination that allows for a passed pawn (a pawn with no opposing pawns in front of it.) \\
\tt{pawnPromotion} & There exists a combination that allows for a pawn promotion. \\
\tt{perpetualCheck} & There exists a combination that forces a perpetual check/draw. \\
\tt{pin} & There exists a combination that allows for a pin. \\
\tt{pullingPiecesToBadSquares} & There exists a combination that forces an opposing piece to move to a bad square. \\
\tt{queenCapture} & There exists a combination to capture the opposing Queen. \\
\tt{quietMates} & There exists a quiet move (one that looks inconsequential on the surface) leading to mate. \\
\tt{removalOfDefence} & There exists a maneuver to remove an opponent’s defender. \\
\tt{unpinning} & There exists a maneuver to unpin a piece. \\
\tt{wrongBishop} & The opponent’s only bishop is limited in use because of the pawn structure of the game. \\ \hline
\end{tabular}
\caption{List of high-level themes we attempt to automatically identify, along with explanations}
\label{table:themes}
\end{table*}

@@

\section{Feature Extraction}
\subsection{Sparse auto-encoder}
The sparse auto-encoder is one of many methods in use for unsupervised feature learning~\cite{Coates}. It takes the form of a feedforward neural network, which consists of an input layer of N nodes, some number of hidden layers, and output layer of M nodes. The value of each layer $\vec{l_i}$ is determined with a weight matrix $W_i$ and a bias vector $b_i$: \[L_i = f(W_i \cdot \vec{l_{i-1}} + b_i),\] where $f(\cdot)$ is a non-linearity like the sigmoid $f(x) = \frac{1}{1 + e^{-z}}$ applied component-wise.

@@

\section{Experimental methodology}
We processed all of the labeled training data to obtain a mapping from each unique game prefix (for most of the experiments, just the initial specified configuration of the board in the labeled game) to the labels that apply to it; most of the prefixes only had one label, while others had two or three. We then segmented the data into three disjoint sets: 90\% for training, 5\% as the validation set (for use with tuning our classifiers), and 5\% as the test set. The segmenting was done by random selection from the set of all labeled games.

@@

\subsection{Support vector machine classifier}
We note that any number of labels may apply to some test input that we want to classify; a straightforward use of a binary (or even multiclass) classifier, as provided by most implementations of support vector machines, will not suffice as it can only assign one label. Therefore, we take the one-versus-all strategy of training many binary SVMs; for each label that we consider, we train a binary classifier with all training games that have the given label as positive examples, and all other games as negative examples. We used SVMlight~\cite{svmlight}, a popular third-party implementation of SVMs, for our experiments, using the manually-generated features extracted from each labeled data point.

@@

\subsection{Random baseline}
To use as a baseline and ensure that our classifiers are giving sensible results, we randomly made predictions for every test game, where each one assigned each possible label at random, with the possibility of the label being given based on the proportion of the training games which have that label. The random predictions were made 1,000 times, with the predictions made independently at random, and the error rates were averaged.

\section{Experimental results}
\subsection{SVM classifier}
\begin{table*}
\centering
\begin{tabular}{lccccccc}
\hline
\textbf{Type} & \textbf{TP} & \textbf{FP} & \textbf{TN} & \textbf{FN} & \textbf{Precision} & \textbf{Recall} & \textbf{F1} \\ \hline
Unsegmented & 0.21\% & 0.00\% & 96.89\% & 2.90\% & 100\% & 6.71\% & 12.57\% \\
Baseline (Random) & 0.42\% & 2.76\% & 94.12\% & 2.69\% & 13.32\% & 13.65\% & 13.48\% \\
\hline
Segmented & 0.13\% & 0.02\% & 98.34\% & 1.51\% & 86.67\% & 7.65\% & 14.05\% \\
Baseline (Random) & 0.12\% & 1.49\% & 96.87\% & 1.51\% & 7.65\% & 7.55\% & 7.60\% \\
\end{tabular} 
\caption{Classification results for the games that have not been segmented by the color of the active player, and those that have been. In the former, precision clearly outperforms a random baseline, but recall falls short, resulting in a lower F1 score. In the latter, precision has fallen relative to the unsegmented version, but recall has risen, resulting in a slightly higher F1 score. In contrast, the random model became much worse in both ways.}
\label{table:svmResults}
\end{table*}

@@

When we extended the labeling to also incorporate the first mover's color information, the SVM model's recall was improved at a minor cost to precision, resulting in a higher F1 score. In contrast, the random predictions became much worse: Both the precision and recall scores fell as a result of this change, which reduced the evenness of the label distribution among the training games, and reduced the amount of games with each label (since games are now spread over twice as many tags).
\begin{table}
\centering
\begin{tabular}{lr}
\hline
\textbf{Tag} & \textbf{Error rate} \\ \hline
\tt{queenCapture}  & 0.6452\% \\
\tt{kingInCenter}  & 0.6452\% \\
\tt{perpetualCheck}  & 0.6452\% \\
\tt{drawCombinationsStalemate}  & 3.8710\% \\
\tt{bishopPair}  & 0.6452\% \\
\tt{doubleAttack}  & 0.6452\% \\
\tt{clearance}  & 2.5806\% \\
\tt{openingClosingDiagonals}  & 0.6452\% \\
\tt{openingClosingFiles}  & 0.6452\% \\
\tt{enclosedKings}  & 1.2903\% \\
\tt{discoveredAttack}  & 2.5806\% \\
\tt{passedPawn}  & 1.2903\% \\
\tt{quietMates}  & 3.8710\% \\
\tt{pin}  & 5.1613\% \\
\tt{removalOfDefence}  & 22.5806\% \\
\tt{deflection}  & 25.1613\% \\
\tt{decoy}  & 9.6774\% \\
\tt{interference}  & 3.8710\% \\
\tt{backrankWeakness}  & 3.2258\% \\
\tt{freeingBlockingSquares}  & 1.9355\% \\
\tt{pawnPromotion}  & 6.4516\% \\
\tt{pullingPiecesToBadSquares}  & 0.6452\%
\end{tabular} 
\caption{Percent of incorrect classification (false negatives and false positives) for various tags.}
\label{table:negatives}
\end{table}

Given the high precision but low recall, we delved deeper to analyze the sources of errors and true positives. As we show in Table~\ref{table:negatives}, we found that {\tt deflection} and {\tt removalOfDefence} are much more likely to cause false negatives than other tags. In addition, we found that all of the true positives are for the {\tt quietMates} tag. 9 of the 11 true positive {\tt quietMates} games were classified correctly using only the features that describe the contents of the board. On the other hand, 5 true positive {\\tt quietMates} games were correctly classified when using only features about the possible moves in that position. Interestingly, disabling the ``checkmate move'' feature did not affect the number of true positives.

\subsection{Classifying with the auto-encoder}
\begin{table*}
\centering
\begin{tabular}{lcccccccc}
\hline
\textbf{Type} & \textbf{Hidden layers} & \textbf{TP} & \textbf{FP} & \textbf{TN} & \textbf{FN} & \textbf{Precision} & \textbf{Recall} & \textbf{F1} \\ \hline
Unsegmented & 0 & 1.49\% & 3.62\% & 93.26\% & 1.61\% & 29.26\% & 48.17\% & 36.40\% \\
Unsegmented & 1 & 1.35\% & 3.83\% & 93.06\% & 1.76\% & 26.01\% & 43.29\% & 32.49\% \\
Unsegmented & 2 & 1.40\% & 4.48\% & 92.41\% & 1.71\% & 23.87\% & 45.12\% & 31.22\% \\
Unsegmented & 3 & 1.40\% & 4.48\% & 92.41\% & 1.71\% & 23.87\% & 45.12\% & 31.22\% \\
\hline
Segmented & 0 & 0.38\% & 1.01\% & 97.35\% & 1.26\% & 27.08\% & 22.94\% & 24.84\% \\
Segmented & 1 & 0.49\% & 2.90\% & 95.46\% & 1.15\% & 14.49\% & 30.00\% & 19.54\% \\
Segmented & 2 & 0.22\% & 1.27\% & 97.09\% & 1.42\% & 14.84\% & 13.53\% & 14.15\% \\
Segmented & 3 & 0.59\% & 3.89\% & 94.47\% & 1.05\% & 13.12\% & 35.88\% & 19.21\% \\
\end{tabular} 
\caption{Auto-encoder classification results for the games that have not been segmented by the color of the active player, and those that have been, as the number of hidden layers varies.}
\label{table:auto-encoder-results}
\end{table*}

@@

These results show higher F1 score than the SVM classifier, with significantly lower precision but also much higher recall. Unfortunately, the results indicate that the representation of the board learned by the auto-encoder does not correlates very well with the concepts we wanted to identify, given that having no hidden layers resulted in higher scores. Adding extra hidden layers seems to not have helped improve the classification results, except in the last instance, where an extra layer improved the F1 score to original levels. It is hard to tell whether some slight change in the auto-encoder training procedure will be able to drastically improve the results, or whether significant changes in the architecture will be required. We also were not able to use the unlabeled chess games we collected with the auto-encoder, due to the long training time required and our relatively slow implementation of its training.

\section{Future Work}
We identify three areas in which future work can make progress on automated sportscasting for chess. It could improve the performance of the auto-encoder so that it can be trained on a larger data set. Furthermore, it could focus on trying different methods to improve the classification quality by leveraging chess-specific tools and features like chess engines and evaluation tools. Moreover, it could investigate an orthogonal path by applying our general approach to other games, including complex ones like StarCraft and simpler ones like Checkers.

\subsection{Performance}
One of the main limitations of our work is performance. Due to how slowly the auto-encoder performs, we were not able to fully exploit the millions of games that we have available to generate features for our unsupervised system. Therefore, extensions to our work should investigate ways to improve the performance so that the auto-encoder can process more inputs and thus generate more relevant feature functions.

\paragraph{Distributed MapReduce.}
Our auto-encoder takes advantage of some of the parallel processing capacity of the host system through a local MapReduce scheme. Extending that with a distributed MapReduce framework like Hadoop may improve performance if the additional cost of network I/O are outweighed by the benefits of parallelization.

\paragraph{Using the GPU.}
Another option that we partially explored is the exploitation of the processing power of the graphics processing unit (GPU). Optimization is a computation-heavy task, and the GPU is specifically designed for massively-parallel floating point computations. We investigated using JavaCL to parallelize the auto-encoder, and demonstrated that it is definitely superior at parallelizable and computation-intensive tasks like matrix multiplication, because the benefits of greater throughput through greater parallelization outweigh the costs of memory transfer.

\subsection{Classification Quality}
Future work should emphasize the classification quality. We demonstrated that general approaches may not be effective enough to be useful for complex games like chess. Therefore, future projects should seek improvements by better taking advantage of the chess-specific tools and libraries that are available like game engines and weight functions. If the classifier incorporated chess-specific features like those generated by chess engines, it could be much better at identifying the types of strategies that are available to the player. For example, a deep traversal of many future paths by a chess engine may reveal that an attack is possible ({\tt attackPossible}) or that the player can force a pawn promotion ({\tt pawnPromotion}). While a purely computational approach to identifying opportunities and strategies may provide higher-quality or more reliable assessments, a statistical approach may be more robust to data sparsity.

One potential improvement that does not involve increased reliance on chess-specific technologies is to implement hierarchical tags so that {\tt matingAttack}, {\tt doubleAttack}, and similar tags imply a more general tag like {\tt attackPossible}. By doing so, there will be a greater balance of positives and negatives, and labels that applied to games but were previously considered wrong are now considered correct. By incorporating hierarchical structure to tags, the system can jointly train and classify games.

Another possible improvement is to use a different classifier than an SVM, such as the perceptron or a maximum entropy classifier. One issue with the SVM is that often the tags are not so clear-cut: what does it mean that an endgame has “good chances”? Rather, there are degrees of how good an endgame is to the player. Therefore, a softer approach like maximum entropy may produce better results.

\subsection{Discussion}
In pursuing this project, we attempted to maintain an approach that is as general as possible so that it can be applied to other contexts, and to minimize the amount of engineering required very specific to chess. However, in attempting to maintain generality, we sacrificed access to powerful but chess-specific tools like chess engines.%, resulting in a classifier that is more surface-level (yet slower) than existing chess analysis tools that leverage the game engine.

A chess-specific rule-based approach should be able to easily identify the more precisely-described themes like {\tt doubleAttack} and {\tt queenCapture} by analyzing the future branches of the game and looking for specific outcomes. However, one advantage of the statistical approach is that feature functions can be written once and reused for all possible themes. In addition, it would help in identifying more subjective themes like {\tt clearance} and {\tt kingToBadSquares} based more on human judgement. In contrast, in order to effectively identify tags using a chess engine, each theme must have a specific heuristic function written for that express purpose to identify certain properties in future states. A better automated sportscasting for chess will leverage the strengths of each approach by using the chess-specific features when they are available, and combining that with results from statistical learning. For instance, while the chess engine can easily identify imminent mate combinations, it would require a coordination of both the engine and the statistical learning tool to identify threshold functions like {\tt kingToBadSquares}. Machine learning would be used to optimize weights and thresholds (e.g. for whether a king’s position is bad) while the engine would be used to explore future paths.

Based on the performance of our general approach, we believe that there is an inherent trade off between how general the approach is and the quality of the outcomes.

\subsection{Generalization to Other Games}
One goal of this project was to develop a statistical machine learning approach that was general enough to be easily adapted to other applications like StarCraft or Checkers. A future work can explore this possibility by simply creating features for other applications.

@@

\section{Acknowledgements}
We would like to thank Dan Klein for teaching us statistical machine learning techniques, and Soda Hall for accommodating us for many nights.

>>>>

\documentclass[11pt]{article}
\usepackage{acl-hlt2011}
\usepackage{color}
\usepackage{graphicx}
\usepackage{times}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{algorithmic}
\usepackage{multirow}
\usepackage{url}
\DeclareMathOperator*{\argmax}{arg\,max}
\setlength\titlebox{6.5cm}    % Expanding the titlebox
\frenchspacing

@@

\begin{abstract}
We investigate applications of statistical learning to automated 
sportscasting for chess. In particular, this project studies two methods 
to automatically identify high level structures in chess games using 
support vector machines. Our first classifier is based on manual feature 
selection, while our second classifier employs sparse auto-encoding to 
automatically generate a set of feature vectors for a chess game. We 
then compare the results of the two different approaches and showed 
improvements using auto-encoding techniques. Finally, we conclude with 
results, analysis of errors, and ideas for potential future work.
\end{abstract}

@@

Unfortunately, we found that our approaches did not deliver accurate 
results in our experiments. In particular, the auto-encoder's output did 
not help significantly as a useful criterion for distinguishing between 
different types of themes. We examine the types of correct predictions 
made by our classifiers and discuss how we may improve our results in 
the future.

@@

\paragraph{Deep learning.} These techniques attempt to capture complex 
relationships between input variables in order to transform the input 
into a different representation which better expresses the relevant 
structures that make up the input. Specifically, they compose many 
layers of non-linearities in order to compactly represent complex 
functions which can model these relationships. They have been used 
successfully for many different types of tasks, especially in computer 
vision~\cite{Vincent,Lee,Socher}.

@@

However, we identified some shortcomings of the AMBL system. First, it 
relies on experts to provide arguments as to why a certain position 
exemplifies (or doesn.t) a particular positional feature. While this is 
an improvement from defining a classification rule manually, it is 
infeasible to repeat this procedure multiple times. Second, the 
arguments that the experts are allowed to provide rely on a preset list 
of features.

@@

\begin{table*}[p]
\centering
\vspace{-0.7cm}
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{Name} & \textbf{Description} \\ \hline
\tt{attackPossible} & The current player has opportunity to attack. \\
\tt{backrankWeakness} & There exists a future checkmate by a rook or 
queen on the edges of the board. \\
\tt{bishopPair} & There exists a decisive strategic advantage for the 
player with two bishops. \\
\tt{clearance} & There exists a maneuver to break the pawn structure. \\ 
\tt{closure} & There exists a maneuver to fix the pawn structure. \\
\tt{decoy} & There exists a maneuver to distract a crucial piece. \\
\tt{defensePressure} & There exists a maneuver to break the opposing 
King.s defense. \\
\tt{deflection} & There exists a maneuver to deflect an attack.  \\ 
\tt{discoveredAttack} & There exists a maneuver to allow for a 
discovered attack (an attack that was blocked by a friendly piece.) \\
\tt{doubleAttack} & There exists a move that simultaneously attacks at 
least two crucial squares. \\
\tt{drawCombinationsStalemate} & There exists a combination that allows 
for decisive material advantage or immediate checkmate. \\
\tt{enclosedKings} & There exists a maneuver to trap the opposing King. 
\\
\tt{freeingBlockingSquares} & There exists a move to free a crucial 
square. \\
\tt{goodEndgameChances} & There exists a move to allow for a favorable 
endgame. \\
\tt{inBetweenMove} & There exists an interrupting move that allows for 
decisive material advantage of immediate checkmate. \\ 
\tt{interference} & There exists a maneuver to block a crucial piece. \\ 
\tt{interruption} & There exists a maneuver that interrupts the 
opponent.s tactics. \\
\tt{kingInCenter} & There exists a decisive strategic advantage for the 
player with the King in the center of the board. \\
\tt{kingToBadSquares} & There exists a combination that forces the 
opposing King to move to a bad square. \\
\tt{longDiagonals} & There exists a strategic advantage for the player 
that controls squares on the long diagonal. \\
\tt{matingAttack} & There exists a decisive mating attack. \\
\tt{openingClosingDiagonals} & There exists a maneuver to clear a 
diagonal. \\
\tt{openingClosingFiles} & There exists a maneuver to clear a file. \\
\tt{overload} & There exists a combination that makes a defending piece 
defend too many squares at once. \\
\tt{passedPawn} & There exists a combination that allows for a passed 
pawn (a pawn with no opposing pawns in front of it.) \\
\tt{pawnPromotion} & There exists a combination that allows for a pawn 
promotion. \\
\tt{perpetualCheck} & There exists a combination that forces a perpetual 
check/draw. \\
\tt{pin} & There exists a combination that allows for a pin. \\
\tt{pullingPiecesToBadSquares} & There exists a combination that forces 
an opposing piece to move to a bad square. \\
\tt{queenCapture} & There exists a combination to capture the opposing 
Queen. \\
\tt{quietMates} & There exists a quiet move (one that looks 
inconsequential on the surface) leading to mate. \\
\tt{removalOfDefence} & There exists a maneuver to remove an opponent.s 
defender. \\
\tt{unpinning} & There exists a maneuver to unpin a piece. \\
\tt{wrongBishop} & The opponent.s only bishop is limited in use because 
of the pawn structure of the game. \\ \hline
\end{tabular}
\caption{List of high-level themes we attempt to automatically identify, 
along with explanations}
\label{table:themes}
\end{table*}

@@

\section{Feature Extraction}
\subsection{Sparse auto-encoder}
The sparse auto-encoder is one of many methods in use for unsupervised 
feature learning~\cite{Coates}. It takes the form of a feedforward 
neural network, which consists of an input layer of N nodes, some number 
of hidden layers, and output layer of M nodes. The value of each layer 
$\mathbf{l_i}$ is determined with a weight matrix $W_i$ and a bias 
vector $b_i$: \[\mathbf{l_i} = f(W_i \cdot \mathbf{l_{i-1}} + b_i),\] 
where $f(\cdot)$ is a non-linearity like the sigmoid $f(x) = \frac{1}{1 
+ e^{-z}}$ applied component-wise.

@@

\section{Experimental methodology}
We processed all of the labeled training data to obtain a mapping from 
each unique game prefix (for most of the experiments, just the initial 
specified configuration of the board in the labeled game) to the labels 
that apply to it; most of the prefixes only had one label, while others 
had two or three. We then segmented the data (consisting of about 3000 
board configurations) into three disjoint sets: 90\% for training, 5\% 
as the validation set (for use with tuning our classifiers), and 5\% as 
the test set. The segmenting was done by random selection from the set 
of all labeled games.

@@

\subsection{Support vector machine classifier}
We note that any number of labels may apply to some test input that we 
want to classify; a straightforward use of a binary (or even multiclass) 
classifier, as provided by most implementations of support vector 
machines, will not suffice as it can only assign one label. Therefore, 
we take the one-versus-all strategy of training many binary SVMs; for 
each label that we consider, we train a binary classifier with all 
training games that have the given label as positive examples, and all 
other games as negative examples. We used {\tt 
SVMlight}~\cite{svmlight}, a popular third-party implementation of SVMs, 
for our experiments, using the manually-generated features extracted 
from each labeled data point.

@@

\subsection{Random baseline}
To use as a baseline and ensure that our classifiers are giving sensible 
results, we randomly made predictions for every test game, where each 
one assigned each possible label at random, with the possibility of the 
label being given based on the prior probability of the label, or the 
proportion of the training games which have that label. The random 
predictions were made 1,000 times, with the predictions made 
independently at random, and the error rates were averaged.

\section{Experimental results}
\subsection{SVM classifier}
\begin{table*}
\centering
\begin{tabular}{lccccccc}
\hline
\textbf{Type} & \textbf{TP} & \textbf{FP} & \textbf{TN} & \textbf{FN} & 
\textbf{Precision} & \textbf{Recall} & \textbf{F1} \\ \hline
Unsegmented & 0.21\% & 0.00\% & 96.89\% & 2.90\% & 100\% & 6.71\% & 
12.57\% \\
Baseline (Random) & 0.42\% & 2.76\% & 94.12\% & 2.69\% & 13.32\% & 
13.65\% & 13.48\% \\
\hline
Segmented & 0.13\% & 0.02\% & 98.34\% & 1.51\% & 86.67\% & 7.65\% & 
14.05\% \\
Baseline (Random) & 0.12\% & 1.49\% & 96.87\% & 1.51\% & 7.65\% & 7.55\% 
& 7.60\% \\ \hline
\end{tabular} 
\caption{Classification results for the games that have not been 
segmented by the color of the active player, and those that have been. 
In the former, precision clearly outperforms a random baseline, but 
recall falls short, resulting in a lower F1 score. In the latter, 
precision has fallen relative to the unsegmented version, but recall has 
risen, resulting in a slightly higher F1 score. In contrast, the random 
model became much worse in both ways.}
\label{table:svmResults}
\end{table*}

@@

When we extended the labeling to also incorporate the first mover's 
color information, the SVM model's recall was improved at a minor cost 
to precision, resulting in a higher F1 score. In contrast, the random 
predictions became much worse: Both the precision and recall scores fell 
as a result of this change, which reduced the evenness of the label 
distribution among the training games, and reduced the amount of games 
with each label (since games are now spread over twice as many tags).
\begin{table}
\centering
\begin{tabular}{lr}
\hline
\textbf{Tag} & \textbf{Error rate} \\ \hline
\tt{queenCapture}  & 0.6452\% \\
\tt{kingInCenter}  & 0.6452\% \\
\tt{perpetualCheck}  & 0.6452\% \\
\tt{drawCombinationsStalemate}  & 3.8710\% \\
\tt{bishopPair}  & 0.6452\% \\
\tt{doubleAttack}  & 0.6452\% \\
\tt{clearance}  & 2.5806\% \\
\tt{openingClosingDiagonals}  & 0.6452\% \\
\tt{openingClosingFiles}  & 0.6452\% \\
\tt{enclosedKings}  & 1.2903\% \\
\tt{discoveredAttack}  & 2.5806\% \\
\tt{passedPawn}  & 1.2903\% \\
\tt{quietMates}  & 3.8710\% \\
\tt{pin}  & 5.1613\% \\
\tt{removalOfDefence}  & 22.5806\% \\
\tt{deflection}  & 25.1613\% \\
\tt{decoy}  & 9.6774\% \\
\tt{interference}  & 3.8710\% \\
\tt{backrankWeakness}  & 3.2258\% \\
\tt{freeingBlockingSquares}  & 1.9355\% \\
\tt{pawnPromotion}  & 6.4516\% \\
\tt{pullingPiecesToBadSquares}  & 0.6452\% \\ \hline
\end{tabular} 
\caption{Percent of incorrect classification (false negatives and false 
positives) for various tags.}
\label{table:negatives}
\end{table}

Given the high precision but low recall, we delved deeper to analyze the 
sources of errors and true positives. As we show in 
Table~\ref{table:negatives}, we found that {\tt deflection} and {\tt 
removalOfDefence} are much more likely to cause false negatives than 
other tags. In addition, we found that all of the true positives are for 
the {\tt quietMates} tag. 9 of the 11 true positive {\tt quietMates} 
games were classified correctly using only the features that describe 
the contents of the board. On the other hand, 5 true positive {\tt 
quietMates} games were correctly classified when using only features 
about the possible moves in that position. Interestingly, disabling the 
``checkmate move'' feature did not affect the number of true positives.

\subsection{Classifying with the auto-encoder}
\begin{table*}
\centering
\begin{tabular}{ccccccccc}
\hline
\textbf{Type} & \textbf{Hidden layers} & \textbf{TP} & \textbf{FP} & 
\textbf{TN} & \textbf{FN} & \textbf{Precision} & \textbf{Recall} & 
\textbf{F1} \\ \hline
Unsegmented & 0 & 1.49\% & 3.62\% & 93.26\% & 1.61\% & 29.26\% & 48.17\% 
& 36.40\% \\
Unsegmented & 1 & 1.35\% & 3.83\% & 93.06\% & 1.76\% & 26.01\% & 43.29\% 
& 32.49\% \\
Unsegmented & 2 & 1.40\% & 4.48\% & 92.41\% & 1.71\% & 23.87\% & 45.12\% 
& 31.22\% \\
Unsegmented & 3 & 1.40\% & 4.48\% & 92.41\% & 1.71\% & 23.87\% & 45.12\% 
& 31.22\% \\
\hline
Segmented & 0 & 0.38\% & 1.01\% & 97.35\% & 1.26\% & 27.08\% & 22.94\% & 
24.84\% \\
Segmented & 1 & 0.49\% & 2.90\% & 95.46\% & 1.15\% & 14.49\% & 30.00\% & 
19.54\% \\
Segmented & 2 & 0.22\% & 1.27\% & 97.09\% & 1.42\% & 14.84\% & 13.53\% & 
14.15\% \\
Segmented & 3 & 0.59\% & 3.89\% & 94.47\% & 1.05\% & 13.12\% & 35.88\% & 
19.21\% \\ \hline
\end{tabular} 
\caption{Auto-encoder classification results for the games that have not 
been segmented by the color of the active player, and those that have 
been, as the number of hidden layers varies.}
\label{table:auto-encoder-results}
\end{table*}

@@

These results show higher F1 score than the SVM classifier, with 
significantly lower precision but also much higher recall. 
Unfortunately, the results indicate that the representation of the board 
learned by the auto-encoder does not correlate very well with the 
concepts we wanted to identify, given that having no hidden layers 
resulted in higher scores. Adding extra hidden layers seems to not have 
helped improve the classification results, except in the last instance, 
where an extra layer improved the F1 score to original levels. It is 
hard to tell whether some slight change in the auto-encoder training 
procedure will be able to drastically improve the results, or whether 
significant changes in the architecture will be required. We also were 
not able to use the unlabeled chess games we collected with the 
auto-encoder, due to the long training time required and our relatively 
slow implementation of its training.

\subsection{Analysis}
Given the nature of labeled training data and their assembly from 
disparate sources, there exists some redundancy among the labels, and 
games may not have all the labels that they should have in our 
chess-description ontology. Furthermore, many of the labels had very few 
(fewer than twenty) games as positive examples, and performance may 
improve greatly if we had a larger training set to use.

\section{Future Work}
We identify three areas in which future work can make progress on 
automated sportscasting. Using a higher-performance auto-encoder 
training procedure and training it on a larger data set could help 
improve the results. We could also extend our system with an automated 
text-generation layer. In addition, leveraging chess-specific tools and 
features like chess engines and evaluation tools could help improve the 
quality of the output. Alternatively, an orthogonal path of applying 
similar general approaches to other games and attempting to learn 
higher-level constructs, including complex ones like StarCraft and 
simpler ones like Checkers, merits exploration.

\subsection{Performance}
One of the main limitations of our work is performance. Due to the slow 
performance of the auto-encoder, we were not able to fully exploit the 
millions of games that we have available to generate features for our 
unsupervised system. Therefore, extensions to our work should 
investigate ways to improve the performance so that the auto-encoder can 
process more inputs and thus generate more relevant feature functions.

\paragraph{Distributed training.}
Our auto-encoder takes advantage of some of the parallel processing 
capacity of the host system through a local MapReduce scheme. Extending 
that with, for example, a distributed MapReduce framework like Hadoop 
may improve performance if the additional cost of network I/O are 
outweighed by the benefits of parallelization.

\paragraph{Using the GPU.}
Another option that we partially explored is the exploitation of the 
processing power of the graphics processing unit (GPU). Function 
optimization, as used in training the auto-encoder, is a 
computation-heavy task, and the GPU is specifically designed for 
massively-parallel floating point computations. We investigated using 
JavaCL, an interface to using the GPU for general purpose computation, 
to parallelize the auto-encoder, and demonstrated that it is definitely 
superior at parallelizable and computation-intensive tasks like matrix 
multiplication, because the benefits of greater throughput through 
greater parallelization outweigh the costs of memory transfer.

\subsection{Text Generation and Linguistic Grounding}
One further avenue of exploration is a system that automatically 
generates text given a chess game with tagged positions. The naive 
approach would be to build a template-based text generation system, 
which could be as simple as generating a formatted sentence highlighting 
hand-selected salient features for each tag or something more 
complicated such as a context-free grammar. In the context of this 
project, we found that some chess positions were positive examples of 
multiple tags. As such, another possible avenue of research could be to 
conglomerate the annotations for each specific position and generate an 
annotation that makes sense in its context.

While we did use high-level structures as defined by experts and 
discussed in detailed commentary of chess games, we did not attempt to 
learn and mimic the actual structures used in the natural language 
commentary. To have a useful commentating system for an intricate game 
like chess, we need to both discover the high-level structures in the 
game, as well as know how to talk about the high-level structures. 
Extending the approach taken by some of the automated sportscasting work 
discussed in the Related Works section, it may be possible to learn both 
high-level concepts that are present in a game, and how to realize 
descriptions of them in language.

\subsection{Improving Results}
In pursuing this project, we attempted to maintain an approach that is 
as general as possible so that it can be applied to other contexts, and 
to minimize the amount of engineering required for components that are 
very specific to chess. Given the complexity of chess, a relatively 
straightforward, but chess-specific, way to improve the performance 
would be to incorporate tools like chess engines. Computer chess has 
been an active field of research, and much work has gone into game 
engines and weight functions which excel at the game.

For example, a chess-specific rule-based approach should be able to 
easily identify the more precisely-described themes like {\tt 
doubleAttack} and {\tt queenCapture} by analyzing the future branches of 
the game and looking for specific outcomes. However, this approach would 
require that each theme must have a specific heuristic function written 
for that express purpose to identify certain properties in future states 
in order to effectively identify tags using a chess engine. In addition, 
it may be difficult to identify more subjective themes like {\tt 
clearance} and {\tt kingToBadSquares} that are based more on human 
judgement.

This suggests a combined approach which can leverage the strengths of 
each approach by using the chess-specific features when they are 
available, and combining that with results from statistical learning. 
For instance, while the chess engine can easily identify imminent mate 
combinations, it would require a coordination of both the engine and the 
statistical learning tool to identify threshold functions like {\tt 
kingToBadSquares}. Machine learning would be used to optimize weights 
and thresholds (e.g. for whether a king's position is bad) while the 
engine would be used to explore future paths. For example, a deep 
traversal of many future paths by a chess engine may reveal that an 
attack is possible ({\tt attackPossible}) or that the player can force a 
pawn promotion ({\tt pawnPromotion}).

\subsection{Generalization to Other Games}
One goal of this project was to develop a statistical machine learning 
approach that was general enough to be easily adapted to other 
applications like StarCraft or Checkers. Given the light dependence on 
chess-specific concepts of the approach described in this paper, 
specifically with the auto-encoder, we should be able to easily adapt 
this to other turn-based games.

@@

\section{Acknowledgements}
We would like to thank Dan Klein for teaching us statistical machine 
learning techniques, Darren Kuo for helping with initial planning and 
game de-duplication, and Soda Hall for accommodating us for many days 
and nights.
